\LARGE Untangling the link between traits, size and growth rate in plants

### 2016-09-16 ms/ms.tex
Towards a full draft

\tableofcontents
% Consider Ecol Letters "Ideas and Perspectives": maximum of 7500 words (main text), and 10 figures, tables or boxes.

@@

@@

Functional traits capture core differences in the strategies plants use to generate and invest surplus energy \citep{Wright-2004, Chave-2009, Westoby-2002}. Although most plants have the same basic physiological function and key resource requirements (carbon, nitrogen and water), species differ considerably in rates at which resources are acquired and invested into different tissues. During the last decade, trade-offs related to prominent traits have been quantified and values for some traits -- such as leaf mass per leaf area, wood density, and seed size -- now exist for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. Concomitantly, researchers have been using traits as potential indicators of the growth strategy and ecology of a species.

While the influence of plant traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have only met with mixed success \citep{Wright-2010, Poorter-2008}. In seedlings, the trait leaf mass per leaf area -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly related to relative growth rate in plant mass \citep{Lambers-1992, Wright-2000}. Leaf mass per leaf area and its correlate leaf lifespan were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in trees leaf mass per leaf area was not correlated with diameter growth rate \citep{Poorter-2008, Wright-2010, Herault-2011}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010}, but less so in small plants \citep{Castro-1998}. In summary, traits did not correlate simply and consistently to growth rate.

The challenge of interpreting diverse empirical results seeking to link traits to growth rate is all the harder because while we lack clear theoretical expectations. Generating such expectations is one of the primary roles for theory \citep{Kokko-2007}. A widely-used model shows that for seedlings, mass-based relative growth rate may be linearly and negatively related to leaf mass per leaf area \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of the model suggests a similar relationship may hold at larger sizes \citep{Enquist-2007}. But as described above, this prediction is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers have simply assumed traits to linearly map onto growth rates, i.e. there is a ``fast'' and ``slow'' growth end of the trait spectrum.

A further problem with existing theory is that the effects of traits are all realised via influences on mass production \citep{Enquist-2007}, whereas the influence of some traits such as leaf mass per leaf area and wood density is via allocation to different tissues rather than mass production \citep{Falster-2011, Gibert-2016}. A second issue with theory focussed solely on mass production is that measuring mass production is only practical for small plants that can be easily harvested. For larger plants, growth is more often measured as increment in either diameter or height. For example, natural research centres and forestry agencies have been recording stem diameter growth in 1000's of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As we outline further below, mass production is only one component of plant growth; meaning theory focussing solely on mass production will always be limited.

Ultimately, trait based understanding must be integrated within a more general model of plant growth, also able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Herault-2011}. In contrast, basal-area growth continues to increase with size \citep{Sillett-2010, Stephenson-2014}, due to an increasing influence of stem turnover. And all growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Meanwhile shade tolerance tends to decrease with increasing size \citep{Lusk-2006}, due to increasing costs of maintaining stems \citep{Givnish-1988}, and also vary with traits \citep{Lusk-2006,Poorter-2006}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

In this paper we show how a new mechanistic framework -- called {\plant} \citep{Falster-2016} -- offers fresh insight into the way traits influence plant growth throughout across the life-cycle and explain diverse phenomena (Table \ref{tab:phenomena}). Our primary aim is to explain a pattern that has been gradually emerging and recently solidified: that the effect of traits on plant growth may be modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the sign of the correlation between some traits (such as leaf mass per leaf area and maximum height) and growth rate was modified by size, while for other traits (such as wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, the direction of the correlations and their shifts with size supported predictions generated via a framework describing how traits influence growth. Here we extend the results of taken by \citet{Gibert-2016} to provide a complete, functioning model and explain, physiologically, why it makes sense that the effect of traits changes with size. By disentangling the effects of plant size, light environment and traits on plant growth, we hope to provide a solid theoretical foundation linking trait-based trade-offs to demography and thus provide a platform for understanding growth across diverse species around the world.

@@

We build on the widespread approach used in many vegetation models of explicitly modelling the amounts in different tissues within the plant \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, DeKauwe-2014} (Fig. \ref{fig:conceptual}a). Here we consider the masses $M_i$, areas $A_i$, and diameters $D_i$ of tissues, with the subscript $i$ indicating: l=leaf, b=bark and phloem, s=sapwood, h=heartwood, r=root, a=live (l+b+s+r), t=total (l+b+s+h+r), and st=stem total (s+b+h). The total mass of living tissue is then $M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$, and the standing mass of the plant is $M_{\rm t}=M_{\rm a}+M_{\rm h}$. A full list of variables, units and definitions is given in Table \ref{tab:definitions} and Table \ref{tab:traits}, with further detail on the parameter values applied given in Table \ref{tab:params}.

We assume growth is fundamentally driven by energy income and subsequent distribution throughout the plant. According to a standard model, the amount of biomass available for growth, ${\rm d}B / {\rm d}t$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000} (Fig. \ref{fig:conceptual}a, eqn \ref{eq:dbdt}). Photosynthesis is proportional to the average photosynthetic rate per unit leaf area $\bar{p}$ and total leaf area, $A_{\rm l} = M_{\rm l} / \phi$, where $\phi$ is leaf mass per leaf area. The constant $y$ in eqn \ref{eq:dbdt} converts mol CO$_2$ into units of biomass. We assume that $\bar{p}$ increases with light but with diminishing returns, and respiration and turnover rates of different tissues are constants that might differ by trait (see below). While many vegetation models use a detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature, such detail is not needed here. Even with detailed models, ${\rm d}B / {\rm d}t$ always comes down to simple subtraction of income and losses.

@@

To model growth in either height, leaf area, basal area, standing mass or stem diameter requires that we account not just for net mass production, but also for the costs of building new tissues, allocation to reproduction, and architectural layout. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016}. Relevant equations (eqns \ref{eq:dbdt}-\ref{eq:dDdt}) are given Fig. \ref{fig:conceptual}b. Importantly, most of these terms vary intrinsically with size. The insets in Fig. \ref{fig:conceptual}b show size-related patterns in different variables for a typical plant obtained from applying the allocation model derived below. However, it is important to recognise that eqns \ref{eq:dMtdt}-\ref{eq:dDdt} are mathematically true, derived using the chain rule of differential calculus, and thus must hold true for any class of models where carbon allocation is important.

Note that the growth rate of all size metrics in eqns \ref{eq:dMtdt}-\ref{eq:dDdt} depend on the product of biomass production $\frac{{\rm d}B} {{\rm d}t}$ (from eqn \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining fraction,$1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$, is allocated to reproduction. $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, presumably at 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants. It is important to realise that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even while the net amount of live mass $M_{\rm a}$ does not increase any further.

The growth rate in total standing mass of the plant, $\frac{{\rm d}M_{{\rm t}}}{{\rm d}t}$ is then sum of sapwood turnover (which is replaced by an equivalent amount) and growth in live mass (eqn \ref{eq:dMtdt}).

The remaining growth rates (eqns \ref{eq:daldt}-\ref{eq:dDdt}) all depend on another variable, $\frac{{\rm d}A_{\rm l}} {{\rm d}M_{\rm a}}$, that accounts for the marginal cost of deploying an additional unit of leaf area, including construction of the leaf itself and support tissues of bark, sapwood and roots (eqn \ref{eq:daldmt}). The inverse of this term, $\frac{{\rm d}M_{\rm a} }{{\rm d}A_{\rm l}}$, is the whole plant construction cost per unit leaf area, which can be further decomposed as a sum of tissue-level construction costs per unit of leaf area, with one of these being $\phi = \frac{{\rm d}M_{\rm l} }{{\rm d}A_{\rm l}}$: the trait leaf mass per area.

The rate of height growth (eqn \ref{eq:dhdt}) depends on an additional term, $\frac{{\rm d}H }{{\rm d}A_{\rm l}}$: the growth in plant height per unit growth in leaf area. This variable accounts for the architectural strategy of the plant. Some species tend to leaf out more than grow tall, while other species emphasise vertical extension \citep{Poorter-2006}. Meanwhile, basal area ($A_{\rm st}$) increment can be expressed as the sum of increments in sapwood, bark \& heartwood areas ($A_{\rm s}, A_{\rm b}, A_{\rm h}$ respectively): $\frac{{\rm d}A_{\rm st}}{{\rm d}t}= \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}$. These in turn are related to ratios of sapwood or bark area per leaf area, and sapwood turnover (see eqn \ref{eq:dast}). Finally, diameter growth is then given by the geometric relationship between stem diameter ($D$) and $A_{\rm st}$ (eqn \ref{eq:dDdt}).

\subsection{Shade tolerance}
A useful measure of a plant's shade tolerance is the point at which photosynthetic gains just balance the costs of tissue turnover and respiration \citep{Givnish-1988,Baltzer-2007,Lusk-2013}, sometimes referred to as the the ``whole-plant light compensation point'' (\wplcp{}).  In general, the average assimilation rate per leaf area $\bar{p}$ is expected to increase with light level or canopy openness, $E$. The \wplcp can be estimated by solving the value of $E=E^*$ where $\frac{{\rm d}B} {{\rm d}t}(E)$ from eqn \ref{eq:dbdt} is zero, which occurs when:
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum \, M_{\rm i} \left(k_{\rm i} y + r_{\rm i}\right)}{A_{\rm l}}.
\end{equation}
Here $\bar{p}(E)$ is the rate of photosynthesis at canopy openness $E$. Below we express

@@

As noted above, most of eqns in Fig. \ref{fig:conceptual} arise as mathematical truisms. To make predictions from eqns \ref{eq:dbdt}-\ref{eq:dDdt}, however, requires a specific model of plant construction. The model described below can be considered a first-order functional-balance model. We might also call it ``isometric'', because the assumptions ensure the proportional relationships among the relevant variables are preserved as the plant grows. Table \ref{tab:allometry} provides key equations of the model (see Supplementary material \ref{sec:ff16} or \citet{Falster-2016} for full derivation). In particular, we assume:
\begin{description}
\item[Assumption 1:] A log-log scaling relationship with slope of 0.5 between the leaf area of a plant and its height (eqn \ref{eq:ha}).
\item[Assumptions 2:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of sapwood in a plant (eqn \ref{eq:as}).
\item[Assumptions 3:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of bark and phloem in a plant (eqn \ref{eq:ab}).
\item[Assumptions 4:] A log-log scaling relationship with slope of 1.0 between the leaf area and mass of fine roots in a plant (eqn \ref{eq:mr}).
\item[Assumption 5:] The vertical distribution of leaf area within the canopy of an individual plant remains relatively constant through ontogeny.
\end{description}

Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are implementations of the pipe model \citep{Shinozaki-1964}, whereby the area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant, we follow the model of \citet{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1. Low values result in a conical canopy, as seen in many conifers, while higher values give a top-weighted canopy profile similar to those seen among angiosperms.

The functional-balance assumptions are well-supported by data, at least to first approximation. Fig. \ref{fig:assumptions} shows data sourced from the Biomass and Allometry Database (\textsc{baad}) \citep{Falster-2015b}, which includes records for 21084 individuals on 656 species for various size metrics of individual plants. Dashed lines indicate the slope of predictions under functional-balance assumptions in Table \ref{tab:allometry}a. As is evident, the lines fitted within each species roughly align with predicted functional-balance assumption Fig. \ref{fig:assumptions}a-c.

Combined, these assumptions in Table \ref{tab:allometry}a lead directly to equations describing the mass of sapwood and bark in relation to leaf, and the amount of leaf in relation to height (Table \ref{tab:allometry}b). Substituting from Table \ref{tab:allometry} into eqns \ref{eq:dhdt}, \ref{eq:dast}, \ref{eq:dDdt} then gives a the necessary components in Fig. \ref{fig:conceptual} and thereby an explicit model for plant growth in relation to size. Note also that no assumptions are made about the diameter and total mass of the stem in relation to other components, e.g. height: these relationships arise as emergent properties, via integrating stem turnover (eqns \ref{eq:ha}- \ref{eq:hm}).

The most important prediction arising from the eqns in Table \ref{tab:allometry} is that the amount of live tissue supporting each unit of leaf area increases with plant height. We can derive an explicit prediction in relation to the assumed allocation model:
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Note that $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are traits, i.e. properties that are assumed approximately constant through ontogeny, so that their product is a constant. Again, the predicted relationship is supported by available data from the \textsc{baad} (Fig. \ref{fig:assumptions}d).

@@

This new growth model captures the intrinsically size-dependent nature of plant growth, recovering well-known empirical patterns (Table \ref{tab:phenomena}). Height growth shows a hump-shaped pattern with size, first increasing then decreasing. This pattern results from systematic changes in the four components of eqn \ref{eq:dhdt} with size (Fig. \ref{fig:conceptual}), including a strong decline in the fraction of plant that is leaf declines with increasing size (Fig. \ref{fig:conceptual}), and increasing reproductive allocation. In contrast, basal-area growth continues to increase with size \citep{Sillett-2010, Stephenson-2014}, due to an increasing influence of stem turnover. Diameter growth shows a weakly hump shaped curve \citep{Herault-2011}, tapering off slightly at larger sizes, in part because of allometric effects (eqn \ref{eq:dDdt}, but mainly because of increased reproductive allocation in older trees (Fig?). All growth measures decrease sharply with size when expressed as relative growth rates \citep{Iida-2014}.

\subsection{How traits influence growth rate}

Distinguish 3 types of trait.

Growth traits

For growth traits can derive rule:
\begin{equation}\label{eq:dg4}
\frac{\frac{\partial \left( \frac{ {\rm d}B} {{\rm d}t}\right)}{\partial x}} {\frac{ {\rm d}B} {{\rm d}t}} < \frac{ \frac{\partial \left(\frac{{\rm d}M_{\rm a}} {{\rm d}A_{\rm l}}\right)}{\partial x}}{\frac{{\rm d}M_{\rm a}} {{\rm d}A_{\rm l}}}.
\end{equation}
In words, there will be a negative relationship between trait x and growth if the relative change in production due to change in traits is less than the relative change in the total construction cost for a unit of leaf area due to change in trait. Crucially, key elements in eqn \ref{eq:dg3} are modified by height.

>>>>

% \tableofcontents
% Consider Ecol Letters "Ideas and Perspectives": maximum of 7500 words (main text), and 10 figures, tables or boxes.

@@

By disentangling the effects of plant size, light environment and traits on plant growth, we hope to provide a solid theoretical foundation linking trait-based trade-offs to demography and thus provide a platform for understanding growth across diverse species around the world.

@@

In this article we outline how the growth of individual plants is influenced by various functional traits, also the plant's size and light environment. Functional traits capture core differences in the strategies plants use to generate and invest surplus energy \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nitrogen and water), species differ considerably in the rates at which resources are acquired and invested into different tissues. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers have looked to traits as potential indicators of plant growth and demography \citep{Poorter-2008, Wright-2010, Adler-2014}.\footnote{Add references from applied fields, such as invasive species, conservation.}

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly related to relative growth rate in plant mass \citep{Lambers-1992, Wright-2000}. leaf mass per unit leaf area and its correlate leaf lifespan were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, leaf mass per unit leaf area was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. So traits do not correlate simply and consistently to growth rate.

It is now becoming increasingly clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the strength of the correlation between some traits (including leaf mass per unit leaf area and maximum height) and growth rate was modified by size, while for other traits (including as wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, the direction of the correlations and their shifts with size supported predictions generated via a framework describing how traits influence growth.

The challenge of interpreting diverse empirical results seeking to link traits to growth rate was all the harder until the paper by \citet{Gibert-2016}, because we lacked any clear theoretical expectations on why the effect of traits may be moderated by size. Generating appropriate expectations is one of the primary roles for theory \citep{Kokko-2007}. A widely-used model for seedlings suggests a seedling's relative growth rate in mass is linearly and negatively related to the trait leaf mass per unit leaf area \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of the model suggests a similar relationship should hold at larger sizes \citep{Enquist-2007}. But as described above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers expected traits to linearly map onto growth rates, i.e. there is a ``fast'' and ``slow'' growth end for each trait \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One possible reason that theoretical predictions have not been supported is that the effects of traits in existing theory are realised mainly via influences on mass production \citep{Enquist-2007}. By contrast, the physiology of traits such as leaf mass per unit leaf area and wood density suggests they have influence by altering the allocation of energy among different tissues, rather than changing mass production \citep{Falster-2011, Gibert-2016}. A second concern for theory focussed on mass production is that altered mass production is only possible outcome. Measuring mass production is really only practical for small plants that can be easily harvested. For larger plants, growth is more-often measured as increment in either diameter or height. For example, research centres and forestry agencies have been recording stem diameter growth in 1000's of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. Theory focussing on mass production is therefore of limited potential use.

In this paper we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth (Table \ref{tab:phenomena}); and thereby offer fresh insight into the way traits influence plant demography throughout across the life-cycle. This model can be viewed as an integration of several related approaches \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011}. Our primary focus in this article is to explain the pattern that has been gradually emerging: that the effect of traits on plant growth may be modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. As noted above, the paper by \citet{Gibert-2016} outlined a framework describing how traits influence growth. Here we extend the results of \citet{Gibert-2016} to provide a functioning model and explain, from the point of view of energy uptake and allocation, why it makes sense that the effect of some traits changes with size. We also show how our approach can account for other phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment. We submit that, trait-based approaches -- which focus on differences among species -- should seek to be integrated within a general model of plant growth, also able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Herault-2011}. In contrast, basal-area growth continues to increase with size \citep{Sillett-2010, Stephenson-2014}. And all growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Lusk-2008,Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Lusk-2008}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

@@

The {\plant} model builds on the widespread approach used in many vegetation models of explicitly modelling the amounts in different tissues within a plant \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, DeKauwe-2014} (Fig. \ref{fig:conceptual}a). We consider the masses $M_i$, areas $A_i$, and diameters $D_i$ of tissues, where the subscripts indicates tissue type: l=leaf, b=bark and phloem, s=sapwood, h=heartwood, r=root, a=live (l+b+s+r), t=total (l+b+s+h+r), and st=stem total (s+b+h). The total mass of living tissue is then $M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$, and the standing mass of the plant is $M_{\rm t}=M_{\rm a}+M_{\rm h}$. A summary of all variables, units and definitions is given in Table \ref{tab:definitions}-\ref{tab:traits}, with further details on the parameter values applied given in Table \ref{tab:params}.

We assume growth is fundamentally driven by energy income and subsequent distribution throughout the plant. Applying a standard approach, the amount of biomass available for growth, $\frac{{\rm d}B}{{\rm d}}t$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000} (Fig. \ref{fig:conceptual}a, eqn \ref{eq:dbdt}). Photosynthesis is the product of the average photosynthetic rate per unit leaf area ($\bar{p}$) and total leaf area ($A_{\rm l}$). The constant $y$ in eqn \ref{eq:dbdt} converts mol CO$_2$ into units of biomass. We assume that $\bar{p}$ increases with light but with diminishing returns, and respiration and turnover rates of different tissues are constants that might differ with traits (see below). While many vegetation models use a detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature, such detail is not needed here, as even with detailed models, $\frac{{\rm d}B}{{\rm d}}t$ always comes down to simple subtraction of income and losses.

@@

To model growth in either height, leaf area, basal area, standing mass or stem diameter then requires that we account not just for net mass production, but also for the costs of building new tissues, allocation to reproduction, and architectural layout. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016}. Relevant equations are given in Fig. \ref{fig:conceptual}b, (eqns \ref{eq:dbdt}-\ref{eq:dDdt}). As is evident in eqn  Fig. \ref{fig:conceptual}b, the growth rate in plant weight (eqn \ref{eq:dMtdt}), leaf area (eqn \ref{eq:daldt}), height (eqn \ref{eq:dhdt}), stem basal area (eqn \ref{eq:dast}), and stem diameter (eqn \ref{eq:dDdt}) all share some terms in common. Many of the terms in Fig. \ref{fig:conceptual} also vary intrinsically with size. The insets in Fig. \ref{fig:conceptual}b show the size-related patterns in different variables for a typical plant, obtained from applying the specific allocation model in the next section.

The growth rate of all size metrics in eqns \ref{eq:dMtdt}-\ref{eq:dDdt} depend on the product of biomass production $\frac{{\rm d}B} {{\rm d}t}$ (from eqn \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining  $1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$ fraction of mass produced is allocated to reproduction. Presumably, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, close to 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants. Note also that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even if the net amount of live mass $M_{\rm a}$ is not increasing.

The growth rate in the total standing mass of the plant, $\frac{{\rm d}M_{{\rm t}}}{{\rm d}t}$ is then the sum of sapwood turnover (which is replaced by an equivalent amount) and increment in live mass (eqn \ref{eq:dMtdt}).

The remaining growth rates (eqns \ref{eq:daldt}-\ref{eq:dDdt}) all depend on another variable, $\frac{{\rm d}A_{\rm l}} {{\rm d}M_{\rm a}}$, that accounts for the marginal cost of deploying an additional unit of leaf area, including construction of the leaf itself and supporting  bark, sapwood and roots (eqn \ref{eq:daldmt}). The inverse of this term, $\frac{{\rm d}M_{\rm a} }{{\rm d}A_{\rm l}}$, is the whole plant construction cost per unit leaf area, which can be further decomposed as a sum of tissue-level construction costs per unit of leaf area, with one of these being $\phi = \frac{{\rm d}M_{\rm l} }{{\rm d}A_{\rm l}}$: the trait leaf mass per area.

The rate of height growth (eqn \ref{eq:dhdt}) depends on an additional term, $\frac{{\rm d}H }{{\rm d}A_{\rm l}}$: the growth in plant height per unit growth in leaf area. This variable accounts for the architectural strategy of the plant. Some species tend to leaf out more than grow tall, while other species emphasise vertical extension \citep{Poorter-2006}.

The rate of stem-basal-area growth (eqn \ref{eq:dast}) can be expressed as the sum of increments in sapwood, bark and heartwood areas ($A_{\rm s}, A_{\rm b}, A_{\rm h}$ respectively): $\frac{{\rm d}A_{\rm st}}{{\rm d}t}= \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}$. These in turn are related to ratios of sapwood and bark area per leaf area, and sapwood turnover (see eqn \ref{eq:dast}).

Finally, the rate of stem-diameter growth (eqn \ref{eq:dDdt}) is given by a geometric relationship between stem diameter ($D$) and stem area ($A_{\rm st}$). Note that we make no assumptions about the relationship of stem diameter to height or leaf area: these arise as emergent properties, via integrating of stem turnover (eqns \ref{eq:ha}- \ref{eq:htm}).

Eqn \ref{eq:dbdt} can also be used to estimate a measure of shade tolerance, sometimes referred to as the the ``whole-plant-light-compensation point'' ({\wplcp}): the  the light level at which a plant's photosynthetic gains just balance the costs of tissue turnover and respiration \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. In general, average assimilation rate per leaf area $\bar{p}$ is expected to increase with light level or canopy openness, $E$. The {\wplcp} can be estimated by solving the value $E=E^*$ giving $\frac{{\rm d}B} {{\rm d}t}(E)=0$. From eqn \ref{eq:dbdt}, this occurs when
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum_{i={\rm l,b,s,r}} \, M_{\rm i} \left(\frac{k_{\rm i}}{y} + r_{\rm i}\right)}{A_{\rm l}},
\end{equation}
where $\bar{p}(E)$ is the rate of photosynthesis at canopy openness $E$. The {\wplcp} occurs at the points where the photosynthetic production (per unit leaf area) line intersects with the sum of maintenance and respiration costs (per unit leaf area) for each tissue (Fig. \ref{fig:wplcp_idea}). Traits will influence the {\wplcp} if they effect either income or costs. Also, since the amount of stem support naturally increases with height, the {\wplcp} also naturally increases with height \citep{Givnish-1988} (Fig. \ref{fig:wplcp_idea}).

@@

It worth noting that because eqns \ref{eq:dMtdt}-\ref{eq:dDdt} are derived derived using the chain rule of differential calculus, they are mathematically true, and must therefore hold for a wide range of potential growth models. To make explicit predictions via this framework, however, requires a explicit model of plant construction and function; i.e. we must put equations behind all there terms in Fig. \ref{fig:conceptual}b.

Here we adopt a model of plant construction and function that can be considered a first-order functional-balance model, similar to those implemented by \citet{Makela-1997} and \citet{Moorcroft-2001}. We could also call it ``isometric'', because the assumptions preserve the proportions among the relevant variables as the plant grows. Table \ref{tab:allometry} provides key equations of the model (see Supplementary Material \ref{sec:ff16} or \citet{Falster-2016} for full derivation). In particular, we assume:
\begin{description}
\item[Assumption 1:] A log-log scaling relationship with slope of 0.5 between the leaf area of a plant and its height (eqn \ref{eq:ha}).
\item[Assumptions 2:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of sapwood in a plant (eqn \ref{eq:as}).
\item[Assumptions 3:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of bark and phloem in a plant (eqn \ref{eq:ab}).
\item[Assumptions 4:] A log-log scaling relationship with slope of 1.0 between the leaf area and mass of fine roots in a plant (eqn \ref{eq:mr}).
\item[Assumption 5:] The vertical distribution of leaf within the canopy of an individual plant remains constant through ontogeny.
\end{description}
Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are realisations of the pipe model \citep{Shinozaki-1964}, whereby the cross-sectional area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant (assumption 5), we use the model of \citet{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1 (for details, see Supplementary Materials).

Combined, the functional-balance assumptions from Table \ref{tab:allometry}a lead directly to equations describing the mass of sapwood and bark in relation to leaf, and the amount of leaf in relation to height, described in Table \ref{tab:allometry}b. Substituting from Table \ref{tab:allometry} into eqns \ref{eq:dhdt}, \ref{eq:dast}, \ref{eq:dDdt} then gives all the necessary terms needed to implement the growth model described in Fig. \ref{fig:conceptual}.

An important prediction arising from the eqns in Table \ref{tab:allometry} is that the amount of live stem tissue supporting each unit of leaf area increases linearly with  height:
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H,
\end{equation}
where $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are all traits, i.e. properties that are assumed approximately constant through ontogeny.

@@

Our growth model suggests an intrinsically size-dependent pattern of growth, which aligns with well-known empirical patterns (Table \ref{tab:phenomena}). The panels in Fig. \ref{fig:conceptual} show the expected patterns for a typical woody plant. Height growth shows a hump-shaped pattern with size, first increasing then decreasing. This pattern results from systematic changes in the four components of eqn \ref{eq:dhdt} with size (Fig. \ref{fig:conceptual}), including a strong decline in the fraction of plant that is leaf declines with increasing size (Fig. \ref{fig:conceptual}), and increasing reproductive allocation. In contrast, basal-area growth continues to increase with size, due to an increasing influence of stem turnover. Diameter growth shows a weakly hump shaped curve, tapering off slightly at larger sizes, in part because of the allometric conversion from basal area to diameter (eqn \ref{eq:dDdt}, and in part because of increased reproductive allocation in older trees (Fig \ref{fig:conceptual}). All growth measures decrease sharply with size when expressed as relative growth rates.

% Importantly, we can remove these size dependent effects by altering XXX of the assumptions in Table \ref{tab:allometry} (??).
% \footnote{What are key assumptions that induce all size-related effects? My guess is  1) changing sapwood area so that volume scales isometrically with height will do it, and 2) removing reproductive allocation. Add figure in SI?}

\subsection{Changes in height growth rate with trait}
